---
title: "Basics of Dplyr Package in R"
output: html_notebook
---

```{r message=FALSE, warning=FALSE}
# load libraries
library(readr)
library(dplyr)
```

```{r}
# load data frame
artists <- read.csv("artists1.csv")
artists
```

## Piping

One of the features in dplyr package is the %>% pipe function

``` {r}
artists %>% head()

                                        # the two function are equivalent
head(artists)
```
## Colume selection

Through using select(df, col1, col2, ..) function, you'll be able to pick the columns you want to include in a df.
We may use %>% to simplify the readability, 
	```
	df %>% select(col1, col2, ...)
	```

```{r}
# select one column

artist_groups = artists %>%  select(group)


```

```{r}
# select multiple columns

group_info = artists %>% select(group, spotify_monthly_listeners, year_founded)

```

To exclude columns, we use ``` df %>% select(-col1, -col2, ..) ```

``` {r}
# select all columns except a set

no_albums = artists %>% select(-albums)

df_cols_removed = artists %>% select(-genre, -spotify_monthly_listeners, -year_founded)
```

## Filtering Rows with Logic

Through filter function and comparison operators, you'll be able to obatin rows you desire. ```df %>% filter(cond1, cond2, ...)``` 
only returning the rows satisfy con1 **and** cond2,

	In this case, and is represented by ",", or is "|", not is "!"


```{r}
rock_groups = artists %>% filter(genre == "Rock")
rock_groups

popular_rock_groups <- artists %>% filter(genre == "Rock", spotify_monthly_listeners > 20000000)
popular_rock_groups


korea_or_before_2000 <- artists %>% filter(country == "South Korea" | year_founded <= 2000)

not_rock_groups <- artists %>% filter(!(genre == "Rock"))
```

## Arranging Rows

Sorting the rows in ascending order by the column provided as an argument ``` df %>% arrange(col) ```
	For numeric cols, ascending - 1-9
	For character, ascending - A - Z
For descending, ```df %>% arrange(desc(col))```

``` {r}

group_asc <- artists %>% arrange(group)
group_asc

youtube_desc <- artists %>% arrange(desc(youtube_subscribers))

```

## Adding a Column

You may add a new column to the dataset by ```mutate()``` function. You can directly perform arithmetic operation on other columns to get a new column

```{r}

# load new data

dogs <- read_csv("dogs.csv")
head(dogs)

names(dogs)

# add one column
dogs <- dogs %>% mutate(avg_height = (height_low_inches + height_high_inches)/2)


# add multiple column

dogs <- dogs %>% mutate(avg_weight = (weight_low_lbs + weight_high_lbs)/2)

dogs <- dogs %>% mutate(rank_change_13_to_16 = rank_2016-rank_2013)

```

## Transmute Columns

function is used for only returning a dataframe with the new created columns ```transmute()```

``` {r}
dogs <- dogs %>%
  transmute(breed = breed,
            avg_height = (height_low_inches + height_high_inches)/2,
         avg_weight = (weight_low_lbs + weight_high_lbs)/2,
        rank_change_13_to_16 = rank_2016 - rank_2013)

dogs

```

## Rename Columns

Use the function```rename(newname = oldname, ...)```, it enables to rename any cols name, be careful that the old name the variable gets the value.

```{r}

original_col_names <- names(dogs)

original_col_names

dogs <- dogs %>% rename(avg_height_inches = avg_height, avg_weight_lbs = avg_weight,
                        popularity_change_13_to_16 = rank_change_13_to_16)

dogs




```




