---
title: "Introduction to Data Visualization in R Using ggplot2"
output: html_notebook
---

# Grammer of Graphics

1. Data - actual data used
2. Geoms - short for geometries - describe the shapes that represent our data. Whether it be dots on a scatter plot, bar charts on the graph, or a line to the plot of data
3. Aesthetics - visual atriibutes of the plot, including the scales on the axes, the color, the fill, and other attributes concerning appearance


```{r message=FALSE, warning=FALSE}
# load libraries and data
setwd("d:/Data_Scientist/Become_Data_Scientist/R/r_studynotes")
library(readr)
library(dplyr)
library(ggplot2)
movies <- read_csv("movies.csv")
```

```{r}
# Observe layers being added with the + sign
viz <- ggplot(data=movies, aes(x=imdbRating, y=nrOfWins)) +
       geom_point(aes(color=nrOfGenre), alpha=0.5) + 
       labs(title="Movie Ratings Vs Award Wins", subtitle="From IMDB dataset", y="Number of Award Wins", x="Move Rating", color = "Number of Genre")


# Prints the plot
viz
```

## Bound the Data
``` {r}

viz = ggplot(data=movies) # bound the dataset

viz
```

## Add aesthetic

``` {r}
viz = ggplot(data=movies, aes(x=imdbRating, y=nrOfWins)) # bound the dataset
```

## Add Geoms

Create scatter plot ```+ geom_point()```
Create a better visually clear parttern ```+ geom_smooth```

**Note** the function ```ggplot()``` is closed with data and aesthetics, geoms are the extra layers like ```ggploy() + geom_point()```

``` {r}
viz <- ggplot(data=movies, aes(x=imdbRating, y=nrOfWins)) + geom_point() # Create scatter point with geom_point() layer


viz

```
