---
title: "R Syntax And Data Structures"
author: "Peter Hou"
date: "29/04/2020"
output:
  pdf_document: 
  fig_caption: yes
  number_sections: yes
  latex_engine: xelatex
  html_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic Syntax

## Assign values to variables and display

```{r}
x <- "Hello World"

cat(x) ##Display variable

print(x) ##Print variable value
```

``` {r}

y = 5 ## We may also assign variable value with "="
print(5)

```
There is no big difference between "=" and "<-" for variable assignment. Most style guide advocate to use "<-" for the variable assignment. Be consistent with the choice in the program.

As for the technical difference between "=", and "<-", you may try the following code to gain an intuition
``` {r}
matrix(1,nrow=2)
print(nrow)

matrix(1, nrow<-2)
print(nrow)
```

## Important Data Type

- String/Character
- Int
- Double (or Float in other program)
(note "Float and Int" are together called numeric)
- Vector
- Matrix
- Data Frame
- Factor
- ...

``` {r}

y_1 <- "R is interesting" ## This is string
y_2 <- 100 ## This is an integer
y_3 <- 99.5 ## This is a float

# Special note about the class of y_2 and y_3

class(y_2)
class(y_3)

y_4 <- TRUE ## This is boolean, boolean needs to be all capital letters in R
y_5 <- c(1,2,3,4) ## This is a vector
y_6 <- matrix(y_5, nrow = 2, ncol = 2) ## This is a matrix
y_7 <- data.frame(y_6) ## This is a data frame


```

**Outputs:**
```{r echo = FALSE}
print(y_1)
print(y_2)
print(y_3)
print(y_4)

print(y_5)
print(y_6)
print(y_7)
```

### Data Parsing

- Single Value Parsing
``` {r}
n1 <- as.numeric("5.0") ## convert to double, there is no as.flot
typeof(n1)
n2 <- as.integer("5.0")
typeof(n2)

as.character(100)
```

- Vectorized Parsing
`as.` function will return a `vector`

``` {r}
as.numeric(c("2.3", "34", "10.7"))
as.integer(c("2.3", "34", "10.7")) ## note as.integer round down the value which 10.7 becomes 10
as.character(c(2.3 34.0 10.7))
n <- as.character(matrix(1:6, nrow=2)) ## it returns vector instead of a matrix
```

- Data Coercion in `R`
``` {r}
vec1 <- c(1, 2, 3, 4, 5)  ## a vector of all numeric data
class(vec1) ## a vector of all numeric data
vec1[1] <- "hello world" ## changes the first entry of the vector to a string/character
class(vec1) ## a vector of all character data
```
**Data Coercion:** `R` does not want to produce `NA` value if it is not necessary. Thus, will implicitly coerce data in the following order.
- logical < integer < numeric < complex < character < list

## Calculations

- Binary Operators

``` {r}
1+1 ##Addition
1-3 ##Subtraction
3*4 ##Multiplication
3/4 ##Division
2**3 ## Exponential (Type 1)
2^3 ## Exponential (Type 2)
10 %% 3 ## Integer Modulo
```


- Functional Operators

``` {r}

exp(3) #base e exponential
log(exp(3)) #natural log or ln in math context
```
Additionally, there are `sin, cos, tan, abs`...

For example the following code
```{r}
abs((exp(sin(cos(pi)))+2*(-1))/(5/4))
```

# Important Data Type: Deep Dive

## Vector and Vectorized Operation

### Vector Properties

**vector can only store values of the same type**
```{r}
vec_1 <- c(2,3,4,5)
print(vec_1)
```

Otherwise, *Data Coercion* will happen!
```{r}
vec_2 <- c(TRUE, "23", 132)
print(vec_2)
vec_3 <- c(TRUE, 12, FALSE)
print(vec_3)
```

Vector cannot be nested, vector of vector will be converted to single vector by r automatically

``` {r}
vec_4 <- c(23, 10, c(23, 10))
print(vec_4)
```

`R` is for statiscians since real people count from `1`!

Vector Slicing

```{r}
dummy <- c("1", "2", "3", "4", "5", "6")
dummy[3] ## Get the 3rd value from a vector
dummy[2:5] ## Vector slicing, it includes the vector at entry index of 5. Programming language like Python doesn't
dummy[(length(dummy) - 2): length(dummy)] ## last 3 values

vec_1 <- c(2, 3, 4, 5)
vec_1[c(1, 2)] ## vector can also be used for slicing and the entries in the vector are for indexes
```

Append a value to a vector

``` {r}
append(dummy, c("N", "B")) ## Append Method 1
c(dummy, c("N", "B")) ## Append Method 2
dummy # dummy doesn't changes

Remove a value from a vector, but it does not change the original vector, it creates a new vector with the value.

``` {r}
dummy[-length(dummy)] ## Remove last value
dummy
dummy <- dummy[-length(dummy)] ## Need to reassign
``` 

Identify value index in a vector

``` {r}
dummy <- c(3,4,2,1,4,3,6,2,1,3,6)
which(dummy %in% c(6,2) ##output every index that corresponds to value of 6 or 2
which(dummy %in% c(6,2)[1]) ## output only indexes correspond to 6
which(dummy %in% c(6,2)[2]) ## output only indexes correspond to 2
match(c(6,2), dummy) #output the first encountered index for each value
match(c(2,6), dummy)
``` 

Reverse a vector `rev`

``` {r}
dummy <- c(2, 3, 4, 5)
rev(dummy)
```

Sort a vector
```{r}
sort(dummy) #default increasing
sort(dummy, decreasing = TRUE)
```

### Vectorized Operations

We can perform operations on vectors directly with ease!

#### Vector Numerical Data Summaries

``` {r}
set.seed(2020) ## control randomness
dummy <- sample(1:10, 10, replace = TRUE) ## 10 random sample from c(1:10) with replacement
dummy

min(dummy)
median(dummy)
max(dummy)
fivenum(dummy) # Return the five number summary
quantile(dummy, c(0, 0.25, 0.5, 0.75, 1))
``` 
The five number summary in r is for min, lower-hinge value, median, upper-hinge value, max.
The lower-hinge value means the median for the lower half value - every value that is lower than the median including the median. 
The upper-hinge value means similar to lower-hinge which is for the median of the upper half value.

The calculation for quantile in r is 
	p - probabily, n - number of element in the vector
	y = p*(n-1) + 1, then a = floor(y), b = ceiling(y). Let c = min(|b-y|, |a-y|) d = max(|b-y|, |a-y|)
	The resulting value =  vector[c] * (1-p) + vector[d] * p

```{r}
sum(dummy) ## Sum 
mean(dummy) ## Average
sd(dummy) ## Sample Standard Deviation
var(dummy) ## Sample Variance

round(var(dummy), digits = 3) ## Round to 3 decimals
```

$$
\text{Sample Variance}=(\text{Sample SD})^2=\frac{\sum_{i=1}^N(y_i-\bar{y})^2}{N-1}
$$
#### Vector Operations

```{r}
x_vec <- c(1,2,3,4)
y_vec <- c(4,3,2,1)
```

We can think of these two as two $4\times 1$ column vectors. Then,

```{r}
3*x_vec 
3+x_vec
```

$$
3\times
\begin{bmatrix}
1\\
2\\
3\\
4
\end{bmatrix}=\begin{bmatrix}
3\\
6\\
9\\
12
\end{bmatrix},
\hspace{2cm}
\begin{bmatrix}
3\\
3\\
3\\
3
\end{bmatrix}+
\begin{bmatrix}
1\\
2\\
3\\
4
\end{bmatrix}=\begin{bmatrix}
4\\
5\\
6\\
7
\end{bmatrix}
$$

Element-wise addition and multiplication
```{r}
x_vec + y_vec
x_vec * y_vec
```

$$
\begin{bmatrix}
1\\
2\\
3\\
4
\end{bmatrix}+
\begin{bmatrix}
4\\
3\\
2\\
1
\end{bmatrix}=\begin{bmatrix}
5\\
5\\
5\\
5
\end{bmatrix},
\hspace{2cm}
\begin{bmatrix}
1\\
2\\
3\\
4
\end{bmatrix}\odot
\begin{bmatrix}
4\\
3\\
2\\
1
\end{bmatrix}=\begin{bmatrix}
4\\
6\\
6\\
4
\end{bmatrix}
$$

In combination, we can apply any mathematical logic element-wise to a vector efficiently!

```{r}
round(exp(x_vec + y_vec^2), digits = 1)
```

You really should think of a `R` vector like a `Python` list and a `R` list as a `Python` library (hash table).

## List

List can contain any data type without coercion

```{r}
dummy_list <- list("a", 123, TRUE, c(0,1))
dummy_list
```

List can be labelled with a reference

``` {r}
dummy_list <- list(a = 1,
                   b = "b",
                   c = c(0, 1),
                   d = list(4, TRUE))
dummy_list

```
Retrieve value/subset from a list

``` {r}
dummy_list[1] ## This will return a subset list
dummy_list[[1]] ## This will return the actual data
dummy_list$a ## Retrieve using label
dummy_list$d[[1]]
```
This last one can be quite confusing when you are reading a large program, so it is recommended to always label the value with clear names!

``` {r}
dummy_list <- list("Peter", "r")
names(dummy_list) <- list("name", "progamming_language")
str(dummy_list) #alternative way of viewing a list
dummy_list
```

Append an element to a list (actually change it)

```{r}
dummy_list$Hobby <- c("Chess", "Math", "CS")
dummy_list[["Career"]] <- "Actuary"
str(dummy_list)
```

Remove/Override an element from a list
```{r}
dummy_list$name <- NULL
dummy_list[["Career"]] <- "Freelancer"
str(dummy_list)
```

Be very careful about list operations in `R` since mutation can happen. There is a great library that have more extended functionalities related to `R` list called `rlist`. You can look more into that [rlist](https://www.rdocumentation.org/packages/rlist/versions/0.4.6.1).

## Matrix
`R` matrices are like vectors and lists in higher dimensions

``` {r}
## Numerical Matrices

x_matrix <- matrix(c(1:6), nrow=2)
x_matrix

y_matrix <- matrix(c(1:6), nrow = 3, byrow = TRUE)
y_matrix
```

By careful with the different betweent byrow = TRUE or the default byrow = FALSE

``` {r}
## Matrix Tranpose
t(y_matrix)
t(y_matrix) - x_matrix ## Element-wise addition/subtraction

```

``` {r}
num_matrix_x <- matrix(1:9, nrow = 3)
num_matrix_x

num_matrix_y <- matrix(rep(1,9), nrow =3)
num_matrix_y
```

``` {r}
diag(num_matrix_x) ## Diagonal entries
diag(x_matrix)

num_matrix_x * num_matrix_y ## Element-wise multiplication
num_matrix_x %*% num_matrix_y
```

`R` matrix can also contain non-numerical values if you add in a list instead of a vector

``` {r}
dummy_matrix <- matrix(list(a = "1",
                            b = "b",
                            c = 2,
                            d = TRUE), nrow = 2)

dummy_matrix

```

To retrieve value from a matrix
``` {r}
dummy_matrix[1, 2] ## [row, col] notation gives a list

dummy_matrix[1, 2][[1]]  ## This gives the actual value

x_matrix[1, 2] # actual value directly obtained from a numerical matrix
```

## `R` Data Frame

### Data Frame Properties

``` {r}
vec_a  <-  c(1, 2, 3, 4)
vec_b <- c("A", "B", "C", "D")
vec_c <- c(TRUE, FALSE, TRUE, FALSE)
dummy_df <- data.frame(vec_a, vec_b, vec_c)
str(dummy_df)

```

To get a part view of a large dataframe, we could use `head()` or `tail()`

``` {r}

head(dummy_df, 2) # first two rows

tail(dummy_df, 2) # last two rows

```


Rename the columns and rows

``` {r}

colnames(dummy_df) <- c("Index", "Level", "Dog")
str(dummy_df)
```

Retrieve data from a dataframe

``` {r}
dummy_df$Level
dummy_df[2, 1]
dummy_df
```

``` {r}

dummy_df[c("Level", "Dog")] ## Subsetting by columns

```

To add new data into a data frame

``` {r}
Cat <- c(FALSE, TRUE, FALSE, TRUE)
cbind(dummy_df, Cat) ## Add a new column (not changed)
str(dummy_df)
```

``` {r}
new_row <- list(Index = 5, Level = "E", Dog = FALSE)
dummy_df <- rbind(dummy_df, new_row) ## row bind
str(dummy_df)
dummy_df

```

``` {r}
dummy_df$Level <- as.factor(dummy_df$Level)
str(dummy_df)
```

`as.factor` function is used to identify categorical variables which a variable can only be of certain kinds, and there's no clear comparison technique between the kinds. For instance, setting up 1, 2, 3, 4 to be factors meaning 4 is no longer necessary to be considered > 3, and 4, 3 are just symbols now, like employee number.



## Basic Boolean Operations

``` {r}

! TRUE # Not operation

1 & 0 # type 1 and operation, TRUE == 1, FALSE == 0

TRUE && TRUE # type 2 and operation

1 | 0 # type 1 or operation

0 || 0 # type 2 or operation

xor(c(0, 1, 1, 0), c(1, 1, 1, 1)) # elementwise order exclusive-or
```
Note xor outputs TRUE when the two argument differs and FALSE when the two arguments are the same, so `xor(1, 1)` gives FALSE and `xor(0, 0)` also gives FALSE 


## Type checking

```{r}
is.character("Hello!") #Character type check
is.numeric("34.4") #Numeric type check
is.integer(4) #Integer check
is.matrix(dummy_matrix) #Matrix check
is.na(c(NA)) # Missing value check
```

## If statements

General Syntax

```
if (cond1) {
    ...
} else if (cond2) {
    ...
} else {
    ...
} 
```

``` {r}
lst_test <- list("R", 123, TRUE, NA)

if (is.numeric(lst_test[[1]])){
    print(lst_test[[1]])
} else if (is.list(lst_test[2])) {
    str(lst_test[2])
} else {
    print(lst_test[[3]])
}
```

Short-hand `ifelse`

``` {r}
ifelse(is.na(lst_test[[4]]),
       lst_test[[4]] <- 0,
       lst_test[[4]])
str(lst_test)
```

## Looping Structures

Looping in R can be done by iterating through vector, list or matrix

### For loop
``` 
for (val in iterable data type) {
	...
}
```

``` {r}
# Accumulate a vector using val directly

accu <- c()

for (val in 11:15) {
    accu <- c(accu, val)
}

accu
```

``` {r}
# Perform task indirectly using index

for (index in 1:length(accu)){
    print(accu[index])
}
```

``` {r}
# Nested for look for matrix
dummy_matrix

for (row_index in 1:nrow(dummy_matrix)) {
    for (col_index in 1:ncol(dummy_matrix)){
        print(paste("[", as.character(row_index), ",", as.character(col_index), "]"
                    , "is of type", typeof(dummy_matrix[row_index, col_index])))
        }
    }
```

### While Loop

While loop is simply a procedure that will not stop until an exit condition is met

General Syntax

``` 
while (condition) {
	...
}

```


``` {r}
# e.g. half the initial point until it reaches a tolerance

initial_point <- c(0, 1)

while (initial_point[2] > 0.0001) {
    initial_point <- c(initial_point[1] + 1, initial_point[2] / 2)
    }

initial_point[1]
```

## £Ò Functions

General Syntax

``` 
func <- function(para_1, para_2, ..) {
	...
}
```
We can specify the return value using `return(result)`, otherwise, `R` will try to return the last line of your function.

``` {r}
# Mathematical Calculation

euclid_norm <- function(xy_pos) {
    sqrt(xy_pos[1]^2 + xy_pos[2]^2)
}
euclid_norm(c(3,4))
```

``` {r}
# Graphing

plot_over_zero_one <- function(func) {
    x_domain <- seq(from = 0, to = 1, by = 0.01)
    plot(x_domain, func(x_domain))
}

func <- function(x){x/(1+x)}
plot_over_zero_one(func)


```

## Factory Function

A factory function is a function factory that produces function variables.

```{r}
poly_degree_gen <- function(degree){
    function(x){
        x^degree
    }
}
poly_degree_gen(3)(2) ## 2^3
